Synthesizing Symmetric Lenses

### 2019-02-28 papers/SymmetricLenses/main.tex
More comments

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? In particular, what about \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure to a target can throw away information that must
then be restored when transferring from target to source), and even more flexibile
transformations that allow each side to retain information not present in
the other?

@@

The first is pragmatic. ``Classical'' symmetric lenses as presented
by ~\citet{symmetric-lenses} operate
over three structures: the ``left'' structure $X$, the ``right'' $Y$ and a complement
$C$ that contains the information not
present in either $X$ or $Y$. These complements must be stored
and managed somehow.  More fundamentially, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses, which we call \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $X$ or $Y$ when making a round trip. They only need the most
recent instance.

To characterize their expressive power formally, we show that simple
symmetric lenses are exactly symmetric lenses that satisfy an intuitive
property called \emph{forgetfulness}.  We show they are expressive
enough in practice by adding simple symmetric lenses to the Boomerang
language~\cite{boomerang} and applying them to a range of real-world
applications, showing that simple symmetric lenses integrate
smoothly with other advanced lens features provided by Boomerang, including
quotient lenses~\cite{quotientlenses,maina+:quotient-synthesis} and matching
lenses~\cite{matchinglenses}.

This leaves us with just the second difficulty with synthesizing symmetric
lenses: While the number of bijective lenses between two given formats is
typically tiny, the number of simple symmetric lenses is typically enormous.
If a na\"ive search algorithm just selects the first simple symmetric lens
it finds, this will generally \emph{not} be the one the user wanted.  We
need a new principle for identifying ``more likely'' lenses and a more
sophisticated synthesis algorithm that uses this principle to search the
space more intelligently. For these, we turn to information theory.

We consider ``likely'' lenses to be ones that propagate ``a lot'' of information
from the left data format to the right and vice versa. Conversely, unlikely
lenses require a large amount of additional information to recover
either of the formats given the other. For many application scenarios, users
will want to synchronise formats with lots of shared information (and lenses
with little shared information are typically easier to write). Our tool is built
primarily for these complex scnearios, and so lenses that propagate more
information are preferable in our context to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the likelihood of a given lens.

@@

With simple symmetric lenses and our SRE-based likelihood measure in hand, we
develop a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure.
Interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting the source and target formats
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types to find the most likely lens. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, uses a greedy, type-directed
algorithm to find a simple symmetric lens between input SRE pairs, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when a sufficiently likely lens is found.

>>>>

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? Could we even imagine synthesizing all \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure to a target can throw away information that must
then be restored when transferring from target to source), and even more flexibile
transformations that allow each side to retain information not present in
the other?

@@

The first of these is pragmatic. ``Classical'' symmetric lenses as presented
by ~\citet{symmetric-lenses} operate
over three structures: a ``left'' structure $X$, a ``right'' structure
$Y$ and a ``complement'' 
$C$ that contains the information not
present in either $X$ or $Y$. These complements must be stored
and managed somehow.  More fundamentially, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses,  called \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $X$ or $Y$ when making a round trip. They only need the most
recent instance.

Formally, we characterize the expressiveness of simple symmetric lenses by
proving that they are exactly the symmetric lenses
that satisfy an intuitive property called \emph{forgetfulness}.  We also show
they are expressive enough for many practical uses by adding simple
symmetric lenses to 
the Boomerang language~\cite{boomerang} and applying them to a range of
real-world applications.  This exercise also demonstrates that simple
symmetric lenses integrate smoothly with other advanced lens features
provided by Boomerang, including {\em quotient
  lenses}~\cite{quotientlenses,maina+:quotient-synthesis} and {\em matching
  lenses}~\cite{matchinglenses}.\bcp{How does it demonstrate this?  Did we
  actually {\em use} symmetric quotient or matching lenses for anything?  Or
are we just observing that these features can coexist in the same
implementation even if using them together might not make sense?}

This leaves us with the second difficulty in synthesizing symmetric
lenses: Whereas the number of bijective lenses between two given formats is
typically tiny, the number of simple symmetric lenses is typically enormous.
If a na\"ive search algorithm just selects the first simple symmetric lens
it finds, this will generally \emph{not} be the one the user wanted.  We
need a new principle for identifying ``more likely'' lenses and a more
sophisticated synthesis algorithm that uses this principle to search the
space more intelligently.

For these, we turn to information theory. We consider ``likely'' lenses to
be ones that propagate ``a lot'' of information 
from the left data format to the right and vice versa. Conversely, ``unlikely''
lenses are ones that require a large amount of additional information to recover
one of the formats given the other. \bcp{Next two sentences are not very
  persuasive.  And they may be a bit out of date anyway, not that we have
  these annotations.  We can just say that we take ``propagate as much as
  possible'' as the default and use explicit annotations on the source and
  target types to mark information that the user knows should not be propagated.}For many application scenarios, users
will want to synchronise formats with lots of shared information (and lenses
with little shared information are typically easier to write). Our tool is built
primarily for these complex scenarios, and so lenses that propagate more
information are preferable in our context to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the likelihood of a given lens.

@@

With simple symmetric lenses and this SRE-based likelihood measure in hand, we
propose a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure.

Interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting the source and target formats
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types to find the most likely lens. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, uses a greedy, type-directed
algorithm to find a simple symmetric lens between input SRE pairs, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when a sufficiently likely lens is found.
